The Impact of Alternation

### 2017-05-10 draft.tex
ok

  \State let $\Art = \tuple{N,E,\Lambda,\lhd}$ be an ART with root
  node $\rootNode$

@@

  \State let $\SubArt = \tuple{N',E',\Lambda',\lhd'}$ be the subtree
  rooted at $p$ in $\Art$

@@

Formally, an ART $\Art = \tuple{N,E,\Lambda,\lhd}$ is a tree, where
$N$ is a set of nodes, $E \subseteq N \times \Sigma \times N$ is a set
of edges, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a labeling
of the nodes with formulae and $\lhd \subseteq N \times N$ is a set of
covering edges. We require that $(N,E)$ is a directed tree, rooted at
$\rootNode \in N$, where $\Lambda(\rootNode) = \iota$. Then each node
$n \in N$ corresponds to a unique path from the root to $n$, labeled
by a sequence $u$ of input events. The least unfeasible suffix of $u$
is the smallest sequence $v = a_1 \ldots a_k$, such that $u = wv$, for
some $w \in \Sigma^*$ and the formula (\ref{eq:pivot}) below is
unsatisfiable. The \emph{pivot} of $u$ is the node $p$ corresponding
to the start of $v$ in $u$.
\begin{equation}\label{eq:pivot}
  \overline{\pi}(v) \equiv \Lambda(p)[Q_0/Q] \wedge
  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
  \theta_{k+1}(Q_k)
\end{equation}
where $\theta_1,\ldots,\theta_{k+1}$ are defined as in
(\ref{eq:interpolation-problem}), for $\theta_0 \equiv
\Lambda(p)[Q_0/Q]$.

@@

  For the second point, we prove the following invariant: each time
  Algorithm \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$
  of nodes in $\worklist$ contains all the frontier nodes in the ART
  $\tuple{N\cup W,E,\Lambda,\lhd}$ which are not covered by some node
  in $N$, namely that:
  \begin{equation}\label{inv:predabs}
  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
    \not\in E ~\wedge~ (n,m) \not\in \lhd}
  \end{equation}
  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
  = \lhd = \emptyset$. If the invariant holds previously, at line
  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
  executed, because, when the subtree rooted at the pivot $p$ is
  removed, $p$ becomes a member of the set of uncovered frontier
  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
  considers that the invariant holds at line \ref{ln:while} and the
  control follows the else branch at line \ref{ln:else}. In this case,
  the newly created frontier node $s$ is added to $W$ only if it is
  not covered by an existing node in $N$ (line \ref{ln:covered}).

@@

The ART built by Algorithm \ref{alg:impact} corresponds to an
unfolding of the input automaton by enumerating the input event
sequences in $\Sigma^*$. Each time a new node is created, its label is
initially $\top$. This label is subject to subsequent strenghtening,
when a spurious counterexample is encountered. To speedup the
computation of interpolants for spurious counterexamples, we use two
additional labeling functions for bookkeeping. Formally, the ART
structure used by Algorithm \ref{alg:impact} is defined as $\Art =
\tuple{N,E,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and $\lhd$ are
defined as for Algorithm \ref{alg:predabs} and,
moreover: \begin{compactitem}
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
  with time-stamped formulae, and
%
\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
  depth\footnote{We consider that the root node has zero depth.}  $k$
  in the tree $(N,E)$, then we have:
  \[\begin{array}{rcl}
  T(e) & = & \Theta(n,k) \\
  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
  \Theta(n,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
  \end{array}\]
\end{compactitem}
For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
$(n,a,m)$. This additional bookkeeping is used as follows. Given a
sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
the root to a node $n$ in $\Art$, the conjunction:
\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
\rightarrow \bot)\] defines the interpolation problem $\pi(u)$.

@@

  \State let $\Art = \tuple{N,E,\Lambda,R,T,\lhd}$ be an ART with root
  node $\rootNode$

@@

  \State $T \leftarrow T \cup \set{(e, \Theta(n,k))}$

@@

Otherwise, the counterexample is spurious and the procedure enters the
refinement phase (lines \ref{ln:refine-begin} -
\ref{ln:refine-end}). The interpolant for $\mu(n)$ is used to
strenghten the labels of all the ancestors of $n$, by conjoining the
formulae of the interpolant to the existing labels. Then $\Lambda(n)$
becomes equivalent to $\bot$.

@@

Finally, we expand $n$ (lines \ref{ln:expand-begin} -
\ref{ln:expand-end}) by creating a new node for each successor $s$ via
an input event $a \in \Sigma$ and inserting it into the worklist. The
label of the new node is $\top$, its replacement set $R(s)$ is
computed based on $R(n)$ and the label of the edge $(n,a,s)$ is
$\Theta(n,k)$, where $k$ is the level of $n$ in $\Art$.

The correctness of Algorithm \ref{alg:impact} is proved in the
following:

@@

  For the second point, we prove the following invariant: each time
  Algorithm \ref{alg:impact} reaches the line \ref{ln:impact-while},
  the set of nodes $W$ in $\worklist$ contains exactly those nodes in
  the ART $\tuple{N \cup W,E,\Lambda,R,T,\lhd}$ which are
  both\begin{inparaenum}[(i)]
  \item uncovered, and such that
  \item for some $a \in \Sigma$, the edge $(n,a,s)$ is missing from $E$:
  \end{inparaenum}
  \begin{equation}\label{inv:impact}
  W = \set{n \not\in N \mid n \text{ uncovered, } \exists a \in
    \Sigma~ \forall s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus the
  invariant holds trivially. Suppose that (\ref{inv:impact}) holds at
  when reaching line \ref{ln:impact-while} and some node $n$ was
  removed from $W$ and inserted into $N$. We distinguish two cases,
  either: \begin{itemize}
  \item $n$ is covered at line \ref{ln:close-worklist}, in which case
    $W$ remains unchanged, or
  %
  \item $n$ is not covered, in which case for each $a \in \Sigma$ a
    new node $s$ is inserted into $W$ (line \ref{ln:expand-end}) and
    the edge $(n,a,s)$ is inserted into $E$ (line
    \ref{ln:edge-insert}). None of these new nodes are covered and
    none of their successors inserted into $E$.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time line
  \ref{ln:impact-while} is reached, thus it is invariant. If Algorithm
  \ref{alg:impact} returns $\true$ then $W = \emptyset$ and, by
  (\ref{inv:impact}), for each node in $n \in N$ one of the following
  hold: \begin{compactitem}
  \item $n$ is covered, or
  \item for each $a \in \Sigma$ there exists $s \in N$ such that
    $(n,a,s) \in E$. 
  \end{compactitem}
  We prove that, in this case, $\bigvee_{n \in N} \Lambda(n)$ defines
  a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
  \ref{lemma:safety-invariant}. To show that $\bigvee_{n \in N}
  \Lambda(n)$ defines an invariant, let $u \in \Sigma^*$ be an
  arbitrary sequence. \qed}

>>>>

  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,\lhd}$ be an ART

@@

  \State let $\SubArt = \tuple{N',E',p,\Lambda',\lhd'}$ be the subtree
  of $\Art$ rooted at $p$

@@

Formally, an ART $\Art = \tuple{N,E,\rootNode,\Lambda,\lhd}$ is a
tree, where $N$ is a set of nodes, $E \subseteq N \times \Sigma \times
N$ is a set of edges, $\rootNode \in N$ is the root of the directed
tree $(N,E)$, $\Lambda : N \rightarrow \Form^+(Q,\vec{x})$ is a
labeling of the nodes with formulae, such that $\Lambda(\rootNode) =
\iota$ and $\lhd \subseteq N \times N$ is a set of covering
edges. Then each node $n \in N$ corresponds to a unique path from the
root to $n$, labeled by a sequence $u$ of input events. The least
unfeasible suffix of $u$ is the smallest sequence $v = a_1 \ldots
a_k$, such that $u = wv$, for some $w \in \Sigma^*$ and the formula
(\ref{eq:pivot}) below is unsatisfiable. The \emph{pivot} of $u$ is
the node $p$ corresponding to the start of $v$ in $u$.
\begin{equation}\label{eq:pivot}
  \overline{\pi}(v) \equiv \Lambda(p)[Q_0/Q] \wedge
  \theta_1(Q_0 \cup Q_1,\vec{x}_0 \cup \vec{x}_1) \wedge \ldots \wedge
  \theta_{k+1}(Q_k)
\end{equation}
where $\theta_1,\ldots,\theta_{k+1}$ are defined as in
(\ref{eq:interpolation-problem}), for $\theta_0 \equiv
\Lambda(p)[Q_0/Q]$.

@@

  For the second point, we prove the following invariant: each time
  Algorithm \ref{alg:predabs} reaches line \ref{ln:while}, the set $W$
  of nodes in $\worklist$ contains all the frontier nodes in the ART
  $\tuple{N\cup W,E,\rootNode,\Lambda,\lhd}$ which are not covered by some node
  in $N$, namely that:
  \begin{equation}\label{inv:predabs}
  W = \set{n \not\in N \mid \forall m \in N~ \forall a \in \Sigma ~.~ (n,a,m)
    \not\in E ~\wedge~ (n,m) \not\in \lhd}
  \end{equation}
  Initially, this is the case because $W = \set{\rootNode}$ and $N = E
  = \lhd = \emptyset$. If the invariant holds previously, at line
  \ref{ln:while}, it will hold again after line \ref{ln:refine} is
  executed, because, when the subtree rooted at the pivot $p$ is
  removed, $p$ becomes a member of the set of uncovered frontier
  nodes, and is added to $W$ at line \ref{ln:refine}. The last case
  considers that the invariant holds at line \ref{ln:while} and the
  control follows the else branch at line \ref{ln:else}. In this case,
  the newly created frontier node $s$ is added to $W$ only if it is
  not covered by an existing node in $N$ (line \ref{ln:covered}).

@@

The ART built by Algorithm \ref{alg:impact} corresponds to an
unfolding of the input automaton by enumerating the input event
sequences in $\Sigma^*$. Each time a new node is created, its label is
initially $\top$. This label is subject to subsequent strenghtening,
when a spurious counterexample is encountered. To speedup the
computation of interpolants for spurious counterexamples, we use two
additional labeling functions for bookkeeping. Formally, the ART
structure used by Algorithm \ref{alg:impact} is defined as $\Art =
\tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$, where $N,E,\Lambda$ and
$\lhd$ are defined as for Algorithm \ref{alg:predabs} and,
moreover: \begin{compactitem}
\item $R : N \rightarrow 2^Q$ is a labeling of nodes with replacement
  sets, such that $R(\rootNode) = \fv{\Bool}{\iota}$, 
%
\item $T : E \rightarrow \bigcup_{i=0}^\infty
  \Form(Q_i,\vec{x}_i,Q_{i+1},\vec{x}_{i+1})$ is labeling of edges
  with time-stamped formulae, and
%
\item for each edge $e = (n,a,m) \in E$, if $n$ lies at
  depth\footnote{We consider that the root node has zero depth.}  $k$
  in the tree $(N,E)$, then we have:
  \[\begin{array}{rcl}
  T(e) & = & \Theta(n,a,k) \\
  R(m) & = & \bigcup_{q \in R(n)}\fv{\Bool}{\Delta(q,a)} \text{ where} \\
  \Theta(n,a,k) & \equiv & \bigwedge_{q \in R(n)} (q_k \rightarrow
  \Delta(q,a)[Q_{k+1}/Q,\vec{x}_{k}/\overline{\vec{x}},\vec{x}_{k+1}/\vec{x}])
  \end{array}\]
\end{compactitem}
For simplicity, we write $T(n,a,m)$ for $T(e)$, when $e$ is
$(n,a,m)$. This additional bookkeeping is used as follows. Given a
sequence $u = a_1\ldots a_k \in \Sigma^*$ that labels the path from
the root to a node $n$ in $\Art$, the conjunction:
\[\mu(n) \equiv \Lambda(\rootNode) \wedge T(\rootNode,a_1,n_1) \wedge \ldots
T(n_{k-1},a_k,n_k) \wedge \bigwedge_{q \in R(n_k) \setminus F}(q_k
\rightarrow \bot)\] defines the interpolation problem $\pi(u)$.

@@

  \State let $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ be an ART

@@

  \State $T \leftarrow T \cup \set{(e, \Theta(n,a,k))}$

@@

Otherwise, the counterexample is spurious and the procedure enters the
refinement phase (lines
\ref{ln:refine-begin}-\ref{ln:refine-end}). The interpolant for
$\mu(n)$ is used to strenghten the labels of all the ancestors of $n$,
by conjoining the formulae of the interpolant to the existing
labels. Then $\Lambda(n)$ becomes equivalent to $\bot$.

@@

Finally, we expand $n$ (lines
\ref{ln:expand-begin}-\ref{ln:expand-end}) by creating a new node for
each successor $s$ via the event $a \in \Sigma$ and inserting it
into the worklist. The label of the new node is $\top$, its
replacement set $R(s)$ is computed based on $R(n)$ and the label of
the edge $(n,a,s)$ is $\Theta(n,a,k)$, where $k$ is the depth of $n$
in $\Art$.

The following lemma states that Algorithm \ref{alg:impact} builds an
over-approximation of the concrete runs of $\A$.

\begin{lemma}\label{lemma:impact-interpol}
  Given an ART $\Art = \tuple{N,E,\rootNode,\Lambda,R,T,\lhd}$ built
  by Algorithm \ref{alg:impact} and a node $n \in N$, for all $(n,a,m)
  \in E$, we have $\Post{\A}(\Lambda(n),a) \models \Lambda(m)$.
\end{lemma}
\proof{ We distinguish two cases. First, if $(n,a,m)$ occurs on a path
  in $\Art$ that has never been refined, then $\Lambda(m)=\top$ and
  the entailment holds trivially. Otherwise, let $(n_0,a_1,n_1),
  \ldots, (n_{k-1},a_k,n_k)$ be a path, where $n_0 = \rootNode$ and
  $(n,a,m) = (n_{i-1},a_i,n_i)$, for some $i \in [1,k]$ and, moreover,
  $a_1\ldots a_k$ was found, at some point, to be a spurious
  counterexample. Let $\tuple{\top,I_1,\ldots,I_k\equiv\bot}$ be an
  interpolant for $\mu(n_k) \equiv \Lambda(\rootNode) \wedge
  \bigwedge_{i=1}^k \Theta(n_{i-1},a,i-1) \wedge \bigwedge_{q \in R(n_k)}
  (q_k \rightarrow \bot)$. 
\qed}

The correctness of Algorithm \ref{alg:impact} is proved in the
following.

@@

  For the second point, we prove the following invariant: each time
  Algorithm \ref{alg:impact} reaches the line \ref{ln:impact-while},
  the set of nodes $W$ in $\worklist$ contains exactly those nodes in
  the ART $\tuple{N \cup W,E,\Lambda,R,T,\lhd}$ which are
  both\begin{inparaenum}[(i)]
  \item uncovered, and such that
  \item for some $a \in \Sigma$, the edge $(n,a,s)$ is missing from $E$:
  \end{inparaenum}
  \begin{equation}\label{inv:impact}
  W = \set{n \not\in N \mid n \text{ uncovered, } \exists a \in
    \Sigma~ \forall s \in N ~.~ (n,a,s) \not\in E}
  \end{equation}
  Initially, $W = \set{\rootNode}$ and $N=E=\lhd=\emptyset$, thus the
  invariant holds trivially. Suppose that (\ref{inv:impact}) holds at
  when reaching line \ref{ln:impact-while} and some node $n$ was
  removed from $W$ and inserted into $N$. We distinguish two cases,
  either: \begin{itemize}
  \item $n$ is covered at line \ref{ln:close-worklist}, in which case
    $W$ remains unchanged, or
  %
  \item $n$ is not covered, in which case for each $a \in \Sigma$ a
    new node $s$ is inserted into $W$ (line \ref{ln:expand-end}) and
    the edge $(n,a,s)$ is inserted into $E$ (line
    \ref{ln:edge-insert}). None of these new nodes are covered and
    none of their successors inserted into $E$.
  \end{itemize}
  Then the condition (\ref{inv:impact}) holds next time line
  \ref{ln:impact-while} is reached, thus it is invariant. If Algorithm
  \ref{alg:impact} returns $\true$ then $W = \emptyset$ and, by
  (\ref{inv:impact}), for each node in $n \in N$ one of the following
  hold: \begin{compactitem}
  \item $n$ is covered, or
  \item for each $a \in \Sigma$ there exists $s \in N$ such that
    $(n,a,s) \in E$. 
  \end{compactitem}
  We prove that, in this case, $\bigvee_{n \in N} \Lambda(n)$ defines
  a safety invariant and conclude that $L(\A) = \emptyset$, by Lemma
  \ref{lemma:safety-invariant}. To this end, let $u \in \Sigma^*$ be
  an arbitrary sequence and let $v_1$ be the largest prefix of $u$
  that labels a path from $\rootNode$ to some node $n_1 \in N$. 
\qed}
